---
title: "Benchmarking Association Analyses of Continuous Exposures with RNA-seq in Observational Studies: tutorial"
author: "Tamar Sofer & Nuzulul Kurniansyah"
date: "12/29/2020"
output: 
  html_document:
    toc: true
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
Here we demonstrate how to perform association analyses of continuous phenotypes 
with RNA-seq data based on the pipeline proposed in the manuscript ``Benchmarking Association Analyses of Continuous Exposures with RNA-seq in Observational Studies". 

```{r}
base_path <- "~/Documents/GitHub/RNASeq/"
```

# Setting up: packages, functions

## Installation Requirements

- An installed R distribution, such as Microsoft R Open. Use R version 3.6.3 or later.
- RStudio, either the commercial edition or the open-source RStudio Desktop.
- R packages (CRAN)


```{r message=FALSE, warning=FALSE, results='hide'}
list.of.packages <- c("dplyr", "tidyverse","purrr","data.table")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

library(dplyr)
library(tidyverse)
library(purrr)
library(data.table)

```

Install packages from Bioconductor
(we may remove some of these, if we are not doing any comparisons in this code)
```{r message=FALSE, warning=FALSE, paged.print=FALSE, results='hide'}

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

list.of.packages <- c("limma", "qvalue","DESeq2","edgeR")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) BiocManager::install(new.packages)
library(limma)
library(qvalue)
library(DESeq2)
library(edgeR)

```

Source files with functions (this will change once the code is packaged)
```{r include=TRUE, echo=TRUE, eval=TRUE}
files <- list.files(base_path, full.names = TRUE)
for (file in files)(
  source(file)
)
```


# Load example data

-  Load transcripts: A p x n matrix of counts of p transcripts across n individuals. 

```{r}

gene_count_matrix <- fread(paste0(base_path, "Data/GSE156437_RNA_seq.txt"), data.table = F)

row.names(gene_count_matrix) <- gene_count_matrix[,1]

gene_count_matrix <- gene_count_matrix[,-1]

gene_count_matrix <- as.matrix(gene_count_matrix)

storage.mode(gene_count_matrix) <- "integer"
gene_count_matrix[1:10,1:10]
```

- Load phenotype

```{r}
phenotype<- fread(paste0(base_path, "Data/2020-08-28_pheno.csv"), data.table = F)

head(phenotype)
```

- Define the outcome

```{r}
outcome <- "Trait.1"
```


- Define covariates to adjust 

```{r}
covariates <- "Age,BMI,Sex"
```


```{r}
count_matrix_new <- gene_count_matrix[, colnames(gene_count_matrix) %in% phenotype$ID]

stopifnot(all(phenotype$ID %in% colnames(count_matrix_new)))

```
 

## Normalized the transcripts

- The options of normalization method are : median_normalization, SizeFactor, TMM. If normalization method is SizeFactor, you define the phenotype, outcome, and covariates string.


```{r}


median_norm<- normalize_trancript_count(count_matrix_non_zero,covariates_string=NULL, outcome=NULL, phenotype=NULL, normalized_type = "median_normalization")

tmm_norm<- normalize_trancript_count(count_matrix_non_zero,covariates_string=NULL, outcome=NULL, phenotype=NULL, normalized_type = "TMM")

sizefactor_norm<- normalize_trancript_count(count_matrix_non_zero,covariates_string=covariates, outcome=outcome, phenotype=phenotype, normalized_type = "SizeFactor")

```


## Applying filter

- Removing lowly express transcripts.


```{r}

clean_count_matrix<- apply_filters(count_matrix = median_norm, 
                                   median_min = 1, 
                                   expression_sum_min = 10, 
                                   max_min = 10, 
                                   range_min = 5, 
                                   prop_zero_max = 0.5)
```



## Creating residual permutation

- Running residual permutations

```{r}
set.seed(12)
null_pval <- compute_residual_permutation(outcome=outcome,
                          count_matrix=clean_count_matrix,
                          phenotype=phenotype,
                          family="gaussian",
                          covariates_string=covariates_string,
                          gene_IDs=NULL,
                          n_permute=100,
                          log_transform="log_replace_half_min",
                          null_pval=TRUE,
                          null_zcore=FALSE)
head(null_pval)

```


- Running DEG

```{r}

deg <- lm_count_mat(count_matrix=clean_count_matrix, 
                    pheno=phenotype, 
                    trait=outcome, 
                    covariates_string=covariates,
                    gene_IDs=NULL, 
                    log_transform = "log_replace_half_min")
head(deg)
```

## Computing emprical p-value

- Quantile empirical p-value, require null pvalues
- Storey empirical-pvalue, require null zscore

```{r}
quantile_empirical<- quantile_empirical_pvalue(p_values = deg$p_value,  null_p_values= null_pval)

deg_quantile_empirical_pvalue<- cbind(deg, empirical_pvalue=quantile_empirical)
deg_quantile_empirical_pvalue_df<- deg_quantile_empirical_pvalue %>% mutate(empirical_BH=p.adjust(empirical_pvalue, method = "BH"))

head(deg_quantile_empirical_pvalue_df)

head(deg_quantile_empirical_pvalue_df[which(deg_quantile_empirical_pvalue_df$empirical_BH < 0.05),])
```





