---
title: "Benchmarking Association Analyses of Continuous Exposures with RNA-seq in Observational Studies: tutorial"
author: "Tamar Sofer & Nuzulul Kurniansyah"
date: "12/29/2020"
output: 
  html_document:
    toc: true
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
Here we demonstrate how to perform association analyses of continuous phenotypes 
with RNA-seq data based on the pipeline proposed in the manuscript ``Benchmarking Association Analyses of Continuous Exposures with RNA-seq in Observational Studies". 

```{r}
base_path <- "~/Documents/GitHub/RNASeq/"
```

# Setting up: packages, functions

## Installation Requirements

- An installed R distribution, such as Microsoft R Open. Use R version 3.6.3 or later.
- RStudio, either the commercial edition or the open-source RStudio Desktop.
- R packages (CRAN)


```{r message=FALSE, warning=FALSE, results='hide'}
list.of.packages <- c("dplyr", "tidyverse","purrr","data.table")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

library(dplyr)
library(tidyverse)
library(purrr)
library(data.table)

```

Install packages from Bioconductor
(we may remove some of these, if we are not doing any comparisons in this code)
```{r message=FALSE, warning=FALSE, paged.print=FALSE, results='hide'}

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

list.of.packages <- c("limma", "qvalue","DESeq2","edgeR")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) BiocManager::install(new.packages)
library(limma)
library(qvalue)
library(DESeq2)
library(edgeR)

```

Source files with functions (this will change once the code is packaged)
```{r include=TRUE, echo=TRUE, eval=TRUE}
files <- list.files(paste0(base_path, "Code"), full.names = TRUE)
for (file in files)(
  source(file)
)
```


# Load example data
## Load transcripts 
First we load the transcripts, where were obtained from XXXX.
We reformat to a desired form. 

```{r}

rnaseq_matrix <- fread(paste0(base_path, "Data/GSE156437_RNA_seq.txt"), data.table = F)

# make transcript names rownames rather than a column
row.names(rnaseq_matrix) <- rnaseq_matrix[,"ensg"]
rnaseq_matrix <- rnaseq_matrix[,-grep("ensg", colnames(rnaseq_matrix))]

# make a matrix instead of a data.frame
rnaseq_matrix <- as.matrix(rnaseq_matrix)

# take a look 
dim(rnaseq_matrix)
rnaseq_matrix[1:10,1:10]
```
After looking at a few transcript counts, we infer that the data was normalized already, because the counts are not natural numbers. We ignore this fact, force the entries to be integerse, and continue demonstrating the pipeline. 

```{r}
rnaseq_matrix <- round(rnaseq_matrix)
storage.mode(rnaseq_matrix) <- "integer"
```

## Load simulated phenotypes
We simulated in advance a data.frame phenotypes. 

```{r}
phenotype <- fread(paste0(base_path, "Data/2020-08-28_pheno.csv"), data.table = F)
head(phenotype)
```

We define the trait of interest to study as an exposure associated with transcription. It has to be a column name in the phenotype data.frame.

```{r}
trait <- "Trait.1"
```

We will adjust our analysis to the simulated covariates Age, BMI, and Sex. The covriates have to correspond to column names in the phenotype data.frame. In the analysis, we will use a string defining the regression model (just the covariates part of it), so we define it here: 

```{r}
covariates <- "Age,BMI,Sex"
covariates_string <- "Age + BMI + Sex"
```
Note that we can also define the string to be "Age + BMI + as.factor(Sex)", or use interaction terms, like one would use in regression functions in R. 

Match the (simulated) individuals between the phenotype and the RNA-seq count matrix. Make sure the IDs overlap.
```{r}
IDs_both <- intersect(phenotype$ID, colnames(rnaseq_matrix))
rnaseq_matrix <- rnaseq_matrix[, IDs_both] 
phenotype <- phenotype[match(IDs_both, phenotype$ID),]
```
 

# Normalize the RNA-seq dataset
The options of normalization method are: median_normalization, SizeFactor, TMM. 
If the selected normalization method is SizeFactor, the users needs to define the phenotype, outcome, and covariates string.


```{r}


median_norm <- normalize_trancript_count(rnaseq_matrix,
                                          normalization_type = "median_normalization")

tmm_norm <- normalize_trancript_count(rnaseq_matrix,
                                      normalization_type = "TMM")

sizefactor_norm <- normalize_trancript_count(rnaseq_matrix,
                                             phenotype=phenotype, 
                                             covariates_string=covariates_string, 
                                             trait=trait, 
                                             normalization_type = "SizeFactor")

```


## Applying filter

Remove lowly express transcripts.


```{r}
clean_count_matrix <- apply_filters(count_matrix = median_norm, 
                                   median_min = 1, 
                                   expression_sum_min = 10, 
                                   max_min = 10, 
                                   range_min = 5, 
                                   prop_zero_max = 0.5)
```





# Perform differential expression analysis

```{r}

deg <- lm_count_mat(count_matrix=clean_count_matrix, 
                    pheno=phenotype, 
                    trait=trait, 
                    covariates_string=covariates_string,
                    gene_IDs=NULL, 
                    log_transform = "log_replace_half_min")
head(deg)
```

## Apply residual permutations 
We want to generate p-values under the null, in order to compute empirical p-values. Therefore, we create a "residual permuted" trait 100 times, perform differential expression analysis, and use the resulting p-values as our null p-values. 


```{r}
set.seed(12)

## re-do this. 
null_pval <- compute_residual_permutation(outcome=outcome,
                          count_matrix=clean_count_matrix,
                          phenotype=phenotype,
                          family="gaussian",
                          covariates_string=covariates_string,
                          gene_IDs=NULL,
                          n_permute=100,
                          log_transform="log_replace_half_min",
                          null_pval=TRUE,
                          null_zcore=FALSE)
head(null_pval)

```

## Computing emprical p-value

- Quantile empirical p-value, require null pvalues
- Storey empirical-pvalue, require null zscore

```{r}
quantile_empirical<- quantile_empirical_pvalue(p_values = deg$p_value,  null_p_values= null_pval)

deg_quantile_empirical_pvalue<- cbind(deg, empirical_pvalue=quantile_empirical)
deg_quantile_empirical_pvalue_df<- deg_quantile_empirical_pvalue %>% mutate(empirical_BH=p.adjust(empirical_pvalue, method = "BH"))

head(deg_quantile_empirical_pvalue_df)

head(deg_quantile_empirical_pvalue_df[which(deg_quantile_empirical_pvalue_df$empirical_BH < 0.05),])
```





